### 2021-03-26 ch04_papers_ecs_tcr_assessment/papers_ecs_tcr_assessment.tex
Updated license information in every file
Further improved text on reasons for high ECS in CMIP6

%=============================================================================%
%                            Doctoral Dissertation                            %
%                               Manuel Schlund                                %
%                                  (c) 2021                                   %
%=============================================================================%
%                        Papers ECS and TCR assessment                        %
%=============================================================================%

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\LargeFigureWidth{}]{
    ch04_papers_ecs_tcr_assessment/figs/cloud_phase_change_feedback.pdf}
  \caption[
    Schematic illustration of the strong negative shortwave cloud phase change
    feedback over the Southern Ocean.
  ]{
    Schematic illustration of the strong negative shortwave cloud phase change
    feedback over the Southern Ocean, which is present in earlier \acs{CMIP}
    model generations like \acs{CMIP}5 (green arrows). Due to substantial
    improvements in the microphysical representation of mixed-phase clouds in
    many \acs{CMIP}6 models, the predominantly ice clouds over the Southern
    Ocean in the present-day climate of previous climate model generations have
    been replaced with clouds that predominantly consist of (supercooled)
    liquid clouds \autocite{BodasSalcedo2019, Gettelman2019}. Thus, in the
    affected \acs{CMIP}6 models, this cloud phase change due to warming is
    reduced by allowing for supercooled cloud liquid, which substantially
    reduces the strong negative cloud phase change feedback (orange arrows).
    This leads to an increase in the global net cloud feedback
    \autocite{BodasSalcedo2019, Tan2016} and the \acf{ECS} in the corresponding
    \acs{CMIP}6 models \autocite{Andrews2019, Gettelman2019}, which is a
    possible explanation for the high \acs{ECS} values found in many
    \acs{CMIP}6 models.
  }
  \label{fig:04:cloud_phase_change_feedback}
\end{figure}

However, it is well known that the low-level mixed-phase clouds over the
Southern Ocean in these earlier climate models are biased towards too large
amounts of ice crystals and too little amounts of supercooled liquid water when
compared to satellite observations \autocite{BodasSalcedo2016}. Thus, the size
of the resulting strong negative cloud phase change feedback in these models
has long been questioned \autocite{McCoy2015, Tan2016}. Improvements in the
representation of mixed-phase clouds are known to reduce this long-standing
cloud phase bias over the Southern Ocean \autocite{BodasSalcedo2016, McCoy2016}
and to improve the representation of both cloud microphysical structure and
cloud radiative impacts in this region \autocite{Hyder2018, Kay2016}. For this
reason, the microphysical representation of mixed-phase clouds has been updated
in many \acs{CMIP}6 models. For example, in \acs{CESM}2, a new mixed-phase ice
nucleation scheme by \textcite{Hoose2010} replaces the \textcite{Meyers1992}
empirical scheme used in predecessor models \autocite{Gettelman2019}.
Simulations with the \textcite{Meyers1992} scheme show almost no supercooled
liquid water in high-latitude clouds and a strong negative cloud phase feedback
when ice turns to liquid. On the contrary, simulations with the new ice
nucleation scheme by \textcite{Hoose2010} show a strong reduction in this
feedback. Since the new scheme also reduces significant cloud phase biases over
the Southern Ocean that are present in the old scheme, the reduced negative
cloud phase feedback is thought to be more consistent with observations of the
current climate \autocite{Gettelman2019}. Another example is the \acs{HadGEM}3
model, for which a new cloud scheme that accounts for turbulent production of
liquid water in mixed-phase clouds has been implemented
\autocite{BodasSalcedo2019}. The new scheme parameterizes the role of
subgrid-scale turbulence in the production and maintenance of supercooled
liquid water \autocite{Furtado2016}. As a consequence, the cloud liquid water
path over the Southern Ocean is increased, which is more in line with
observations from microwave satellite retrievals. Moreover, the increase in the
cloud liquid water path with warming is reduced in \acs{HadGEM}3, leading to a
reduction in the negative cloud phase change feedback over the Southern Ocean
\autocite{BodasSalcedo2019}. In addition, \textcite{Williams2020} show that
short-term forecasts created with \acs{HadGEM}3 are more accurate than
comparable simulations from predecessor model versions (using traditional
metrics from numerical weather prediction), which indicates an improvement of
the physical processes in the new climate model.

Overall, these changes in many of the \acs{CMIP}6 models lead to a better
agreement of the \acs{CMIP}6 \ac{MMM} of the \ac{TOA} shortwave \ac{CRE} with
corresponding observations \autocite{Bock2020}. Moreover, the strong negative
cloud feedback described above that results from a cloud phase change from ice
clouds in the present-day to liquid clouds in the future is substantially
reduced in those \acs{CMIP}6 models that simulate predominantly liquid clouds
over the present-day Southern Ocean. This is illustrated in
\cref{fig:04:cloud_phase_change_feedback} (orange arrows). As a consequence,
this reduction of the negative shortwave cloud phase change feedback increases
the global net cloud feedback \autocite{BodasSalcedo2019, Tan2016} and
eventually the \ac{ECS} of the affected \acs{CMIP}6 models
\autocite{Andrews2019, Gettelman2019}, offering a possible explanation for the
high \ac{ECS} values found in many \acs{CMIP}6 models.

>>>>

%=============================================================================%
%                            Doctoral Dissertation                            %
%                            (c) by Manuel Schlund                            %
%=============================================================================%

%=============================================================================%
% This work is licensed under a
% Creative Commons Attribution 4.0 International License.
%
% You should have received a copy of the license along with this
% work. If not, see <http://creativecommons.org/licenses/by/4.0/>.
%=============================================================================%

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\LargeFigureWidth{}]{
    ch04_papers_ecs_tcr_assessment/figs/cloud_phase_change_feedback.pdf}
  \caption[
    Schematic illustration of the strong negative shortwave cloud phase change
    feedback over the Southern Ocean.
  ]{
    Schematic illustration of the strong negative shortwave cloud phase change
    feedback over the Southern Ocean, which is present in earlier \acs{CMIP}
    model generations like \acs{CMIP}5 (green arrows). Due to improvements in
    the microphysical representation of mixed-phase clouds, some \acs{CMIP}6
    models simulate a higher fraction of (supercooled) liquid water in the
    clouds over the present-day Southern Ocean than their corresponding
    predecessor models \autocite{BodasSalcedo2019, Gettelman2019}. Thus, in the
    affected \acs{CMIP}6 models, the strong negative cloud feedback that
    results from a cloud phase change from ice clouds in the present-day to
    liquid clouds in the future is substantially reduced (orange arrows). This
    leads to an increase in the global net cloud feedback
    \autocite{BodasSalcedo2019, Tan2016} and the \acf{ECS}
    \autocite{Andrews2019, Gettelman2019}, which is a possible explanation for
    the high \acs{ECS} values found in some \acs{CMIP}6 models.
  }
  \label{fig:04:cloud_phase_change_feedback}
\end{figure}

However, it is well known that the low-level mixed-phase clouds over the
Southern Ocean in these earlier climate models are biased towards too large
amounts of ice crystals and too little amounts of supercooled liquid water when
compared to satellite observations \autocite{BodasSalcedo2016}. Thus, the size
of the resulting strong negative cloud phase change feedback in these models
has long been questioned \autocite{McCoy2015, Tan2016}. Improvements in the
representation of mixed-phase clouds are known to reduce this long-standing
cloud phase bias over the Southern Ocean \autocite{BodasSalcedo2016, McCoy2016}
and to improve the representation of both cloud microphysical structure and
cloud radiative impacts in this region \autocite{Hyder2018, Kay2016}. For this
reason, the microphysical representation of mixed-phase clouds has been updated
in some \acs{CMIP}6 models. For example, in \acs{CESM}2, a new mixed-phase ice
nucleation scheme by \textcite{Hoose2010} replaces the \textcite{Meyers1992}
empirical scheme used in predecessor models \autocite{Gettelman2019}.
Simulations with the \textcite{Meyers1992} scheme show almost no supercooled
liquid water in high-latitude clouds and a strong negative cloud phase change
feedback when ice turns to liquid. On the contrary, simulations with the new
ice nucleation scheme by \textcite{Hoose2010} show a strong reduction in this
feedback. Since the new scheme also reduces significant cloud phase biases over
the Southern Ocean that are present in the old scheme, the reduced negative
cloud phase change feedback is thought to be more consistent with observations
of the current climate \autocite{Gettelman2019}. Another example is the
\acs{HadGEM}3 model, for which a new cloud scheme that accounts for turbulent
production of liquid water in mixed-phase clouds has been implemented
\autocite{BodasSalcedo2019}. The new scheme parameterizes the role of
subgrid-scale turbulence in the production and maintenance of supercooled
liquid water \autocite{Furtado2016}. As a consequence, the cloud liquid water
path over the Southern Ocean is increased, which is more in line with
observations from microwave satellite retrievals. Moreover, the increase in the
cloud liquid water path with warming is reduced in \acs{HadGEM}3, leading to a
reduction in the negative cloud phase change feedback over the Southern Ocean
\autocite{BodasSalcedo2019}. In addition, \textcite{Williams2020} show that
short-term forecasts created with \acs{HadGEM}3 are more accurate than
comparable simulations from predecessor model versions (using traditional
metrics from numerical weather prediction), which indicates an improvement of
the physical processes in the new climate model.

Overall, such changes in the \acs{CMIP}6 models contribute to a better
agreement of the \acs{CMIP}6 \ac{MMM} of the \ac{TOA} shortwave \ac{CRE} with
observations resulting in higher average pattern correlations of the model
output with satellite observations \autocite{Bock2020}. Moreover, the strong
negative cloud feedback described above that results from a cloud phase change
from ice clouds in the present-day to liquid clouds in the future is
substantially reduced in those \acs{CMIP}6 models that simulate a higher
fraction of (supercooled) liquid water in the clouds over the present-day
Southern Ocean than their corresponding \acs{CMIP}5 predecessors. This is
illustrated in \cref{fig:04:cloud_phase_change_feedback} (orange arrows). The
reduction of the negative shortwave cloud phase change feedback over the
Southern Ocean increases the global net cloud feedback
\autocite{BodasSalcedo2019, Tan2016} and eventually the \ac{ECS} of the
affected \acs{CMIP}6 models \autocite{Andrews2019, Gettelman2019}, offering a
possible explanation for the high \ac{ECS} values found in some \acs{CMIP}6
models.

### 2021-03-26 ch07_conclusion/conclusion.tex
Updated license information in every file
Further improved text on reasons for high ECS in CMIP6

%=============================================================================%
%                            Doctoral Dissertation                            %
%                               Manuel Schlund                                %
%                                  (c) 2021                                   %
%=============================================================================%
%                                 Conclusion                                  %
%=============================================================================%

@@

In the first study of this thesis, the climate sensitivity metrics \ac{ECS} and
\ac{TCR} are evaluated for the latest generation of \acp{ESM} from \acs{CMIP}6.
This work, which is presented in \cref{ch:04:papers_ecs_tcr_assessment} and
already published in \textcite{Bock2020, Meehl2020}, directly addresses
\cref{enum:01:question_1} (\enquote{\emph{\KeyScienceQuestionOne{}}}). For
\ac{ECS}, a \acs{CMIP}6 model range of $\rangeunit{1.8}{5.6}{K}$ is found,
which is higher on the upper and lower end than any model range from previous
\ac{CMIP} generations before. In comparison to \acs{CMIP}5, the \acs{CMIP}6
\ac{MMM} of \ac{ECS} is about $16 \unit{\%}$ higher ($3.74 \unit{K}$ in
\acs{CMIP}6 versus $3.23 \unit{K}$ in \acs{CMIP}5). Moreover, especially the
upper bound of the \acs{CMIP}6 models range is considerably larger than in the
assessed range of $\rangeunit{1.5}{4.5}{K}$ given by the latest published
\ac{AR} of the \ac{IPCC} from 2013 \autocite{Stocker2013}. The assessed upper
bound of $4.5 \unit{K}$ is exceeded by a third of the \acs{CMIP}6 models, with
many models showing \ac{ECS} values above $5 \unit{K}$. For \ac{TCR}, the model
range of \acs{CMIP}6 is $\rangeunit{1.3}{3.0}{K}$, which also exceeds the
\acs{CMIP}5 range of $\rangeunit{1.1}{2.5}{K}$ and the assessed range from
\acs{AR}5 of $\rangeunit{1.0}{2.5}{K}$. One possible reason for the increased
climate sensitivity in many \acs{CMIP}6 models is the addition of prognostic
aerosol schemes that include aerosol-cloud interactions, which could have
produced overly large negative radiative forcing. To correctly reproduce the
observed historical temperature trend, a stronger response to \ac{GHG} forcing
is required, resulting in a higher climate sensitivity in the corresponding
models. A further reason for the high \ac{ECS} in \acs{CMIP}6 is a change in
the microphysical representation of mixed-phase clouds over the Southern Ocean
in some models. This change was mainly implemented to reduce a long-standing
cloud phase bias over the Southern Ocean that is present in models from
previous \ac{CMIP} generations. Moreover, the \acs{CMIP}6 \ac{MMM} shows an
improved simulation of the shortwave \ac{CRE} when compared to observations.
However, the change in the representation of mixed-phase clouds also
substantially reduces the strong negative shortwave cloud feedback over the
Southern Ocean that is present in previous \ac{CMIP} generations and that
results from a cloud phase change from ice clouds in the present-day to liquid
clouds in the future. In the affected \acs{CMIP}6 models, this cloud phase
change due to warming is reduced since these models simulate less cloud ice
over the present-day Southern Ocean than their predecessor versions with no
supercooled cloud liquid.

>>>>

%=============================================================================%
%                            Doctoral Dissertation                            %
%                            (c) by Manuel Schlund                            %
%=============================================================================%

%=============================================================================%
% This work is licensed under a
% Creative Commons Attribution 4.0 International License.
%
% You should have received a copy of the license along with this
% work. If not, see <http://creativecommons.org/licenses/by/4.0/>.
%=============================================================================%

@@

In the first study of this thesis, the climate sensitivity metrics \ac{ECS} and
\ac{TCR} are evaluated for the latest generation of \acp{ESM} from \acs{CMIP}6.
This work, which is presented in \cref{ch:04:papers_ecs_tcr_assessment} and
already published in \textcite{Bock2020, Meehl2020}, directly addresses
\cref{enum:01:question_1} (\enquote{\emph{\KeyScienceQuestionOne{}}}). For
\ac{ECS}, a \acs{CMIP}6 model range of $\rangeunit{1.8}{5.6}{K}$ is found,
which is higher on the upper and lower end than any model range from previous
\ac{CMIP} generations before. In comparison to \acs{CMIP}5, the \acs{CMIP}6
\ac{MMM} of \ac{ECS} is about $16 \unit{\%}$ higher ($3.74 \unit{K}$ in
\acs{CMIP}6 versus $3.23 \unit{K}$ in \acs{CMIP}5). Moreover, especially the
upper bound of the \acs{CMIP}6 models range is considerably larger than in the
assessed range of $\rangeunit{1.5}{4.5}{K}$ given by the latest published
\ac{AR} of the \ac{IPCC} from 2013 \autocite{Stocker2013}. The assessed upper
bound of $4.5 \unit{K}$ is exceeded by a third of the \acs{CMIP}6 models, with
many models showing \ac{ECS} values above $5 \unit{K}$. For \ac{TCR}, the model
range of \acs{CMIP}6 is $\rangeunit{1.3}{3.0}{K}$, which also exceeds the
\acs{CMIP}5 range of $\rangeunit{1.1}{2.5}{K}$ and the assessed range from
\acs{AR}5 of $\rangeunit{1.0}{2.5}{K}$. One possible reason for the increased
climate sensitivity in some \acs{CMIP}6 models is the addition of prognostic
aerosol schemes that include aerosol-cloud interactions, which might result in
overly large negative radiative forcing. In this case, a stronger response to
\ac{GHG} forcing is required to correctly reproduce the observed historical
temperature trend, resulting in a higher climate sensitivity in the
corresponding models. A further reason for the high \ac{ECS} in \acs{CMIP}6 is
a change in the microphysical representation of mixed-phase clouds in some
\acs{CMIP}6 models. This change was implemented to improve the representation
of the fractioning of cloud ice and cloud liquid in those models, \eg{} by
allowing for (more) supercooled cloud liquid water. In addition, the
\acs{CMIP}6 \ac{MMM} shows an improved simulation of the shortwave \ac{CRE}
when compared to satellite observations. The change in the representation of
mixed-phase clouds in some \acs{CMIP}6 models reduces the strong negative
shortwave cloud feedback over the Southern Ocean that is present in predecessor
versions of these models and that results from a cloud phase change from ice
clouds in the present-day to liquid clouds in the future. In the affected
\acs{CMIP}6 models, this cloud phase change due to warming is reduced since
these models simulate less cloud ice over the present-day Southern Ocean than
their predecessor versions with no supercooled cloud liquid.
